Estimating Body Fat Percentage Using Anthropometric Measurements For Affordable Health Monitoring

### 2024-12-03 other/datasheet/datadictionary.qmd
graph label changes & title fixes

---
title: "Data Dictionary for Body Fat Raw Dataset"
author: 
  - Wendy Yuan
thanks: "Code and data are available at: https://github.com/kiwindyy/Body-Fat"
date: November 29, 2024
date-format: long
format: pdf
---
**Density**

>>>>

---
title: "Data Dictionary for Body Fat Percentage Raw Dataset"
author: 
  - Wendy Yuan
thanks: "Code and data are available at: https://github.com/kiwindyy/Body-Fat"
date: November 29, 2024
date-format: long
format: pdf
---
**Density**

### 2024-12-03 other/datasheet/datasheet.qmd
graph label changes & title fixes

---
title: "Datasheet for Body Fat Dataset"
subtitle: "Predictive Modeling of Body Fat Percentage Using Body Measurements"
author: 
  - Wendy Yuan
thanks: "Code and data are available at: https://github.com/kiwindyy/Body-Fat"
date: November 29, 2024
date-format: long
abstract: "This datasheet describes the Body Fat Dataset from Brigham Young University. It includes measurements like weight, height, and body circumferences from 250 men. The data helps predict body fat percentage using simple methods instead of expensive equipment. It is publicly available but only represents men and lacks diversity."
format: pdf
number-sections: true
bibliography: references.bib
---

>>>>

---
title: "Datasheet for Body Fat Percentage Dataset"
subtitle: "Predictive Modeling of Body Fat Percentage Using Body Measurements"
author: 
  - Wendy Yuan
thanks: "Code and data are available at: https://github.com/kiwindyy/Body-Fat"
date: November 29, 2024
date-format: long
abstract: "This datasheet describes the Body Fat Dataset from Brigham Young University. It includes measurements like weight, height, and body circumferences from 250 men. The data helps predict body fat percentage using simple methods instead of expensive equipment. It is publicly available but only represents men and lacks diversity."
format: pdf
number-sections: true
bibliography: references.bib
---

### 2024-12-03 paper/paper.qmd
graph label changes & title fixes

---
title: "Estimating Body Fat Percentage Using Anthropometric Measurements"
subtitle: "Abdomen as the Strongest Predictor With Contributions From Wrist, Height, and Age For Affordable Health Monitoring"
author: 
  - Wendy Yuan
thanks: "Code and data are available at: https://github.com/kiwindyy/Body-Fat"
date: November 29, 2024
date-format: long
abstract: "This study uses a multi-linear regression model to estimate body fat percentage based on circumference measurements. Abdominal size was the strongest predictor, followed by wrist size, height, and age, which showed minimal contributions. These results demonstrate that accessible methods can effectively estimate body fat, offering methods suitable for non-clinical settings. This approach is important for identifying and managing health risks associated with extremely low or high body fat levels in a practical and affordable way."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

Before analysis, the raw dataset was cleaned to standardize and prepare the data for use. Since the original dataset included measurements in both inches and centimeters, all measurements were converted into centimeters to ensure uniformity. Additionally, measurements in pounds were converted to kilograms. [@tbl-cleaned1] & [@tbl-cleaned2] & [@tbl-cleaned3] show the first 8 observations of the cleaned data. Additional cleaning steps included removing duplicates, addressing outliers, and removing missing observations to improve dataset’s usability. This cleaned data makes sure that the variables in the study are accurate and consistent. For reference, the original dataset before cleaning (first 10 observation) is provided in [@tbl-raw1] & [@tbl-raw2] in @sec-raw, showing the changes made during this process.

@@

# Histogram for Pct.BF
hist(data$Pct.BF, 
     main = "Histogram of Percent Body Fat",
     xlab = "Percent Body Fat",
     col = "#f88379",
     border = "black",
     breaks = 20)
```

@@

# Print the AIC value and final variables
kable(final_variables)
```

@@

# Histogram for each variable
hist(data$Age, main = "Histogram of Age", xlab = "Age", col = "#7dbeff", border = "black")
hist(data$Height, main = "Histogram of Height", xlab = "Height (cm)", col = "#2aebaa", border = "black")
hist(data$Abdomen, main = "Histogram of Abdomen", xlab = "Abdomen (cm)", col = "#eb49a4", border = "black")
hist(data$Wrist, main = "Histogram of Wrist", xlab = "Wrist (cm)", col = "#bb86f7", border = "black")
```

@@

# Generate a pairwise plot
pairwise_plot <- ggpairs(selected_data, 
                         lower = list(continuous = wrap("points", color = "#008080", alpha = 0.4)), 
                         upper = list(continuous = wrap("points", color = "#008080", alpha = 0.4)),
                         diag = list(continuous = wrap("densityDiag", fill = "#008080", alpha = 0.6)))

@@

# Print VIF values
kable(vif_values)
```

@@

# 1. Linearity: Residuals vs Fitted Plot
plot(model$fitted.values, residuals(model), 
     main = "Residuals vs Fitted",
     xlab = "Fitted Values",
     ylab = "Residuals")
abline(h = 0, col = "red")
```

@@

```{r}
#| label: fig-variance
#| fig-cap: Scale Location Verifying Constance Variance Assumption
#| echo: false
#| warning: false
#| message: false

@@

plot(model, which = 3)
```

@@

# 4. Normality: Q-Q Plot of Residuals
qqnorm(residuals(model))
qqline(residuals(model), col = "red")
```

@@

```{r}
#| label: fig-simple
#| fig-cap: Simple Linear Regression Plots Against Body Fat Percent for Age, Height, Abdomen, Wrist
#| echo: false
#| warning: false
#| message: false

p1 <- ggplot(data, aes(x = Age, y = Pct.BF)) +
  geom_point(color = "#7dbeff", alpha = 0.5) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Pct.BF vs Age", x = "Age", y = "Pct.BF") +
  theme_minimal()

p2 <- ggplot(data, aes(x = Height, y = Pct.BF)) +
  geom_point(color = "#2aebaa", alpha = 0.5) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Pct.BF vs Height", x = "Height (inches)", y = "Pct.BF") +
  theme_minimal()

p3 <- ggplot(data, aes(x = Abdomen, y = Pct.BF)) +
  geom_point(color = "#eb49a4", alpha = 0.5) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Pct.BF vs Abdomen", x = "Abdomen (cm)", y = "Pct.BF") +
  theme_minimal()

p4 <- ggplot(data, aes(x = Wrist, y = Pct.BF)) +
  geom_point(color = "#bb86f7", alpha = 0.5) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Pct.BF vs Wrist", x = "Wrist (cm)", y = "Pct.BF") +
  theme_minimal()

@@

# Create the plot
ggplot(coeff_data, aes(x = Estimate, y = Parameter)) +
  geom_point(size = 3, color = "#f88379") +
  geom_errorbarh(aes(xmin = Lower, xmax = Upper), height = 0.2,
                 color = "#f88379", linewidth = 0.8) +
  theme_minimal() +
  labs(
    title = "Model Coefficients",
    subtitle = "90% Confidence Intervals",
    x = "Coefficient Estimate",
    y = "Predictors"
  ) +
  theme(
    axis.title.y = element_text(size = 12, face = "bold"),
    axis.text.y = element_text(size = 10),
    plot.title = element_text(size = 14, face = "bold")
  )

@@

# Split into two smaller tables
part1 <- raw_data[, 1:8]  # First 8 columns
part2 <- raw_data[, 9:16]  # Remaining 8 columns

# Render each part
kable(head(part1, 10), booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

@@

kable(head(part2, 10), booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

@@

```{r}
#| label: fig-fullresidual
#| fig-cap: Residual Plot Verifying the Linearity for All Variables
#| echo: false
#| warning: false
#| message: false

@@

# Linearity: Residuals vs Fitted Plot
plot(fullresidual$fitted.values, residuals(fullresidual), 
     main = "Residuals vs Fitted",
     xlab = "Fitted Values",
     ylab = "Residuals")
abline(h = 0, col = "red")
```

@@

```{r}
#| label: fig-fullvariance
#| fig-cap: Scale Location Verifying Constance Variance for All Variables
#| echo: false
#| warning: false
#| message: false

@@

plot(fullresidual, which = 3)
```

@@

```{r}
#| label: fig-fullnormality
#| fig-cap: QQ-Plot Verifying the Normality of Errors for All Variables
#| echo: false
#| warning: false
#| message: false

@@

# Normality: Q-Q Plot of Residuals
qqnorm(residuals(fullresidual))
qqline(residuals(fullresidual), col = "red")
```

>>>>

---
title: "Estimating Body Fat Percentage Using Anthropometric Measurements For Affordable Health Monitoring"
subtitle: "Abdomen as the Strongest Predictor, Wrist & Height as Moderate Predictors, and Age as the Weakest Predictor"
author: 
  - Wendy Yuan
thanks: "Code and data are available at: https://github.com/kiwindyy/Body-Fat"
date: November 29, 2024
date-format: long
abstract: "This study uses a multi-linear regression model to estimate body fat percentage based on circumference measurements. Abdominal size was the strongest predictor, followed by wrist size, height, and age, which showed minimal contributions. These results demonstrate that accessible methods can effectively estimate body fat, offering methods suitable for non-clinical settings. This approach is important for identifying and managing health risks associated with extremely low or high body fat levels in a practical and affordable way."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

Before analysis, the raw dataset was cleaned to standardize and prepare the data for use. Since the original dataset included measurements in both inches and centimeters, all measurements were converted into centimeters to ensure uniformity. Additionally, measurements in pounds were converted to kilograms. [@tbl-cleaned1] & [@tbl-cleaned2] & [@tbl-cleaned3] show the first 8 observations of the cleaned data. The cleaned data is used for the rest of the analysis in this paper. Additional cleaning steps included removing duplicates, addressing outliers, and removing missing observations to improve dataset’s usability. This cleaned data makes sure that the variables in the study are accurate and consistent. For reference, the original dataset before cleaning (first 10 observation) is provided in [@tbl-raw1] & [@tbl-raw2] in @sec-raw, showing the changes made during this process.

@@

# Histogram for Pct.BF
hist(data$Pct.BF, 
     main = "Histogram of Body Fat Percentage",
     xlab = "Body Fat Percentage",
     ylab = "Frequency",
     col = "#f88379",
     border = "black",
     breaks = 20)
```

@@

# Print the AIC value and final variables
final_variables_table <- data.frame("Selected Variables" = final_variables)
kable(final_variables_table)
```

@@

# Histogram for each variable
hist(data$Age, main = "Histogram of Age", xlab = "Age", ylab = "Frequency", col = "#7dbeff", border = "black")
hist(data$Height, main = "Histogram of Height", xlab = "Height (cm)", ylab = "Frequency", col = "#2aebaa", border = "black")
hist(data$Abdomen, main = "Histogram of Abdomen", xlab = "Abdomen (cm)", ylab = "Frequency", col = "#eb49a4", border = "black")
hist(data$Wrist, main = "Histogram of Wrist", xlab = "Wrist (cm)", ylab = "Frequency", col = "#bb86f7", border = "black")
```

@@

# Generate a pairwise plot
pairwise_plot <- ggpairs(selected_data, 
                         lower = list(continuous = wrap("points", color = "#008080", alpha = 0.4)), 
                         upper = list(continuous = wrap("points", color = "#008080", alpha = 0.4)),
                         diag = list(continuous = wrap("densityDiag", fill = "#008080", alpha = 0.6))) +
  ggtitle("Relationship Between Selected Predictor Variables") +
  theme(plot.title = element_text(hjust = 0.5))

@@

# Convert VIF values to a data frame
vif_table <- data.frame("Predictor Variables" = names(vif_values), 
                        "VIF Values" = as.numeric(vif_values))

# Print the updated VIF table
kable(vif_table)
```

@@

# 1. Linearity: Residuals vs Fitted Plot
plot(model$fitted.values, residuals(model), 
     main = "Residuals vs Fitted Plot",
     xlab = "Fitted Values of Model",
     ylab = "Residuals")
abline(h = 0, col = "red")
```

@@

```{r}
#| label: fig-variance
#| fig-cap: Scale Location Plot Verifying Constance Variance Assumption
#| echo: false
#| warning: false
#| message: false

@@

plot(model, which = 3, sub = "")
```

@@

qqnorm(residuals(model), 
       main = "Q-Q Plot of Residuals", 
       xlab = "Theoretical Quantiles", 
       ylab = "Sample Quantiles")
qqline(residuals(model), col = "red")
```

@@

```{r}
#| label: fig-simple
#| fig-cap: Simple Linear Regression Plots Against Body Fat Percent for Age, Height, Abdomen, Wrist
#| fig-height: 3.5
#| fig-width: 6.2
#| echo: false
#| warning: false
#| message: false

p1 <- ggplot(data, aes(x = Age, y = Pct.BF)) +
  geom_point(color = "#7dbeff", alpha = 0.5) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Age vs Body Fat Percentage", x = "Age (years)", y = "Pct.BF (g/cm^3)") +
  theme_minimal()

p2 <- ggplot(data, aes(x = Height, y = Pct.BF)) +
  geom_point(color = "#2aebaa", alpha = 0.5) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Height vs Body Fat Percentage", x = "Height (cm)", y = "Pct.BF (g/cm^3)") +
  theme_minimal()

p3 <- ggplot(data, aes(x = Abdomen, y = Pct.BF)) +
  geom_point(color = "#eb49a4", alpha = 0.5) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Abdomen vs Body Fat Percentage", x = "Abdomen (cm)", y = "Pct.BF (g/cm^3)") +
  theme_minimal()

p4 <- ggplot(data, aes(x = Wrist, y = Pct.BF)) +
  geom_point(color = "#bb86f7", alpha = 0.5) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Wrist vs Body Fat Percentage", x = "Wrist (cm)", y = "Pct.BF (g/cm^3)") +
  theme_minimal()

@@

# Create the plot
ggplot(coeff_data, aes(x = Estimate, y = Parameter)) +
  geom_point(size = 3, color = "#f88379") +
  geom_errorbarh(aes(xmin = Lower, xmax = Upper), height = 0.2,
                 color = "#f88379", linewidth = 0.8) +
  theme_minimal() +
  labs(
    title = "Body Fat Percentage Model Coefficients",
    subtitle = "90% Confidence Intervals",
    x = "Coefficient Estimate",
    y = "Predictors"
  ) +
  theme(
    axis.title.y = element_text(size = 12, face = "bold"),
    axis.text.y = element_text(size = 10),
    plot.title = element_text(size = 14, face = "bold")
  )

@@

# Split into two smaller tables
raw_part1 <- raw_data[-1, 1:8]

# Update column names to include units
colnames(raw_part1) <- c(
  "Density", "Pct.BF", "Age", "Weight", "Height", "Neck", "Chest", "Abdomen")

# Render each part
kable(head(raw_part1, 10), row.names = FALSE, booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

@@

raw_part2 <- raw_data[-1, 9:16]

colnames(raw_part2) <- c(
  "Waist", "Hip", "Thigh", "Knee", "Ankle", "Bicep", "Forearm", "Wrist")

kable(head(raw_part2, 10), row.names = FALSE, booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

@@

```{r}
#| label: fig-fullresidual
#| fig-cap: Residual Plot Verifying the Linearity for Full Model
#| echo: false
#| warning: false
#| message: false

@@

# Linearity: Residuals vs Fitted Plot
plot(fullresidual$fitted.values, residuals(fullresidual), 
     main = "Residuals vs Fitted Plot",
     xlab = "Fitted Values of Full Model",
     ylab = "Residuals")
abline(h = 0, col = "red")
```

@@

```{r}
#| label: fig-fullvariance
#| fig-cap: Scale Location Plot Verifying Constance Variance for Full Model
#| echo: false
#| warning: false
#| message: false

@@

plot(fullresidual, which = 3, sub = "", ann = FALSE)
title(xlab = "Fitted Values of Full Model", 
      ylab = "sqrt(|Standardized Residuals|)")
```

@@

```{r}
#| label: fig-fullnormality
#| fig-cap: QQ-Plot Verifying the Normality of Errors for Full Model
#| echo: false
#| warning: false
#| message: false

@@

# Normality: Q-Q Plot of Residuals
qqnorm(residuals(fullresidual),
       main = "Q-Q Plot of Residuals for Full Model", 
       xlab = "Theoretical Quantiles of Full Model", 
       ylab = "Sample Quantiles of Full Model")
qqline(residuals(fullresidual), col = "red")
```
